R Notebook

### 2021-06-20 Implementatione.Rmd
nuovi files

---
title: "R Notebook"
output:
  pdf_document: default
  html_notebook: default
---

\textbf{Implementation}

In this section we present some practical illustrations of the algorithm discussed. For the sake of simplicity we use a random walk plus noise model, i.e. the most basic form of a linear Gaussian state-space model. 
\begin{align}
y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \\
x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \\
x_{0} & \sim N(m_{0},C_{0})
\end{align}
As we already know, in this case the filtering distribution can be computed in closed form solutions using the Kalman filter. However, this toy example will be the basis for the implementations of other filtering strategies since we think that it is useful to understand the logic of the algorithms and to compare their performances.
The typical observed process for this kind of model is the one presented in Figure XX. We simulated 500 observation imposing $\sigma^{2}=\tau^{2}=1$.
```{r echo=FALSE, fig.align='center', fig.cap="Toy example: the true states and the observation sequence", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
library(ggplot2)
q025=function(x){quantile(x,0.025)}
q975=function(x){quantile(x,0.975)}

set.seed(1991)
dlm.sim = function(n,sigma,tau,x0){
  x    = rep(0,n)  #state process
  y    = rep(0,n)  #obs process
  x[1] = rnorm(1,x0,tau)
  y[1] = rnorm(1,x[1],sigma)
  for (t in 2:n){
    x[t] = rnorm(1,x[t-1],tau)
    y[t] = rnorm(1,x[t],sigma)
  }
  return(list(x=x,y=y))
}

n=50
tau2=1
sigma2=1
sigma=sqrt(sigma2)
tau=sqrt(tau2)
x0=0

sim<-dlm.sim(n,sigma,tau,x0)
y<-sim$y
x<-sim$x

timeframe<-c(1:length(y))
df<-data.frame(timeframe,y,x)

ggplot(df,aes(x=timeframe))+
  geom_line(aes(y=y, linetype="Observations"))+
  geom_line(aes(y=x, linetype="True States"))+
  scale_linetype_manual(name="",
                      values=c("Observations" = 1,
                               "True States" = 3))+
  labs(x="Time",
       y="")+
  ggtitle("States and Observations")+
  theme_bw()+
  theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
        legend.background = element_rect(fill = "white", colour = "gray30")) +
  theme(plot.title = element_text(hjust = 0.5))
```
The Kalman Filter for this model can be easily implemented. Starting from the filtering distribution at period $t-1$, $x_{t-1}|y_{1:t-1} \sim N(m_{t-1},C_{t-1})$, we compute:
\begin{itemize}
\item the one-step-ahead predictive distribution at time $t-1$
\begin{align*}
x_{t}|y_{1:t-1} & \sim N(a_{t},R_{t})\\
a_{t} & = m_{t-1}\\
R_{t} & = C_{t-1}+\tau^2
\end{align*}
\item the filtering distribution at time $t$ as $p(x_{t}|y_{1:t}) \propto p(x_{t}|y_{1:t-1})p(y_{t}|x_{t})$
\begin{align*}
x_{t}|y_{1:t} & \sim N(m_{t},C_{t}) \\
m_{t} & = \big(1-\frac{R_{t}}{R_{t}+\sigma^2}\big)a_{t}+\frac{R_{t}}{R_{t}+\sigma^2}y_{t} \\
C_{t} & = \frac{R_{t}}{R_{t}+\sigma^2}\sigma^2
\end{align*}
\end{itemize}
```{r}
DLM<-function(data,sig2,tau2,m0,C0){
  n  = length(data)
  m  = rep(0,n)
  C  = rep(0,n)
  for (t in 1:n){
    if (t==1){
      a = m0
      R = C0 + tau2
    }else{
      a = m[t-1]
      R = C[t-1] + tau2
    }
    A = R/(R+sig2)
    m[t] = (1-A)*a + A*y[t]
    C[t] = A*sig2
  }
  return(list(m=m,C=C))
}
```
In the Figure below we show the filtered states states estimated using Kalman Filter with $x_{0} \sim N(0,100)$. The filtered states follow the observations closely and they provide a good approximation of the true states.
```{r echo=FALSE, fig.align='center', fig.cap="Kalman Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
#Kalman Filter
#-------------
m0=0
C0=100
filtval<-DLM(y,sigma2,tau2,m0,C0)
m<-filtval$m
C<-filtval$C
Low = m + qnorm(0.025)*sqrt(C)
Up = m + qnorm(0.975)*sqrt(C)

#Plots (observetions in black and filtered states in red)
ggplot(DLM.df,aes(x=timeframe))+
  geom_line(aes(y=y, col="Observations", linetype="Observations"))+
  geom_line(aes(y=x, col="True States", linetype="True States"))+
  geom_line(aes(y=m, col="Filtered States", linetype="Filtered States"))+
  geom_ribbon(aes(ymin = Low, ymax = Up),
              fill="red",alpha=0.16) +
    scale_color_manual("",
                      values=c("Observations" = "black",
                               "True States" = "black",
                               "Filtered States"= "red"))+
  scale_linetype_manual("",
                      values=c("Observations" = 1,
                               "True States" = 3,
                               "Filtered States"=1))+
  labs(x="Time",
       y="")+
  ggtitle("Kalman Filter")+
  theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
        legend.background = element_rect(fill = "white", colour = "gray30")) +
  theme(plot.title = element_text(hjust = 0.5))
```

\textbf{Implementation}
With reference to the random walk plus noise of section XX, let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \frac{1}{N}\sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. For $t=1,...,n$ where $n$ is the length of the sample, at any iteration
\begin{itemize}
\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
\end{itemize}
```{r}
SISfun<-function(data,N,m0,C0,tau,sigma){
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)
  for(t in 1:length(data)){
    x    = rnorm(N,x,tau)                   #sample from N(x_{t-1},tau)
    w    = w*dnorm(data[t],x,sigma)         #update weight
    xs = rbind(xs,x)
    ws = rbind(ws,w)

@@

```{r echo=FALSE}
N=1000
sis<-SISfun(y,N,m0,C0,tau,sigma)
sisxs<-sis$xs
sisws<-sis$ws
sisess<-sis$ess
sis.data<-data.frame(df,sisxs,sisws,sisess)

ggplot(sis.data,aes(x=timeframe))+
  geom_line(aes(y=sisess))+
    labs(x="Time",
       y="")+
  ggtitle("Effective Sample size")+
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
```

```{r echo=FALSE, fig.align='center', fig.cap="SIS Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
Filtervalues<-function(fun){
  xhat<-sapply(1:n,function(i)
    weighted.mean(fun$xs[i,],fun$ws[i,]))
  sdhat<-sapply(1:n,function(i)
    sqrt(weighted.mean((fun$xs[i,]-xhat[i])^2,fun$ws[i,])))

  return(list(mean=xhat,sd=sdhat))

}

Filtplot<-function(dataframe,fun,title){

Filt<-Filtervalues(fun)
mean<-Filt$mean
sd<-Filt$sd
dataframe<-data.frame(dataframe,mean,sd)

ggplot(dataframe,aes(x=timeframe))+
  geom_line(aes(y=y, col="Observations", linetype="Observations"))+
  geom_line(aes(y=x, col="True States", linetype="True States"))+
  geom_line(aes(y=mean, col="Filtered States", linetype="Filtered States"))+
  geom_ribbon(aes(ymin = mean-1.96*sd, ymax = mean+1.96*sd),
              fill="red",alpha=0.16) +
    scale_color_manual("",
                      values=c("Observations" = "black",
                               "True States" = "black",
                               "Filtered States"= "red"))+
  scale_linetype_manual("",
                      values=c("Observations" = 1,
                               "True States" = 3,
                               "Filtered States"=1))+
  labs(x="Time",
       y="")+
  ggtitle(title)+
  theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
        legend.background = element_rect(fill = "white", colour = "gray30")) +
  theme(plot.title = element_text(hjust = 0.5))
}

Filtplot(df,sis,title="Sequential Importance Sampling Filter")

```
\
\textbf{Particle Filter}
\
With reference to the random walk plus noise of section XX, let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \frac{1}{N}\sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. For $t=1,...,n$ where $n$ is the length of the sample, at any iteration
\begin{itemize}
\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
\end{itemize}
In addition, when $ESS<ESS_{0}$ \footnote{In our example we fix $ESS_{0}=N/2$, this is an arbitrary common rule of thumb.}, resempling applies
\begin{itemize}
\item Draw a sample of size N, $x_{t}^{(1)},...,x_{t}^{(N)}$,from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{itemize}
```{r}
PFfun<-function(data,N,m0,C0,tau,sigma,r){
  if(missing(r)){r=2}else{}
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)

  for(t in 1:length(data)){

    x<-rnorm(N,x,tau)
    w1<-w*dnorm(data[t],x,sigma)

    w = w1/sum(w1)
    ESS  = 1/sum(w^2)

    if(ESS<(N/r)){
      index<-sample(N,size=N,replace=T,prob=w)
      x<-x[index]
      w<-rep(1/N,N)
    }else{}

    xs = rbind(xs,x)
    ws = rbind(ws,w)
    ess =rbind(ess,ESS)
  }
  return(list(xs=xs,ws=ws,ess=ess))
}
```

```{r echo=FALSE, fig.align='center', fig.cap="Effective Sample Size", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
N=1000

pf<-PFfun(y,N,m0,C0,tau,sigma)

ESSplot<-function(dataframe,fun,threshold){
if(missing(threshold)){threshold=T}else{}
xs<-fun$xs
ws<-fun$ws
ess<-fun$ess
dt<-data.frame(dataframe,xs,ws,ess)

ggplot(dt,aes(x=timeframe))+
  geom_line(aes(y=ess))+
  geom_line(aes(y=500),col="orange")+
  labs(x="Time",
       y="")+
  ggtitle("Effective Sample Size")+
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
}
ESSplot(df,pf)

@@

```{r echo=FALSE, fig.align='center', fig.cap="Particle Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
Filtplot(df,pf,"Particle Filter")
```

\textbf{Guided Particle Filter}

Let's consider another toy example. Suppose that we have the following model

```{r}
GPFfun<-function(data,N,m0,C0,tau,sigma,r){
 if(missing(r)){r=2}else{}
   xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)

  for(t in 1:length(data)){

    xprev<-x
    x<-rnorm(N,x,tau)
    w1<-w*dnorm(data[t],x,sigma)*dnorm(x,xprev,tau)*I(x>0)/dnorm(x,xprev,tau)

    w = w1/sum(w1)
    ESS  = 1/sum(w^2)

    if(ESS<(N/r)){
      index<-sample(N,size=N,replace=T,prob=w)
      x<-x[index]
      w<-rep(1/N,N)
    }else{}

    xs = rbind(xs,x)
    ws = rbind(ws,w)
    ess =rbind(ess,ESS)
  }
  return(list(xs=xs,ws=ws,ess=ess))
}
```

\textbf{}

With reference to the linear gaussian model of section XX, a very basic auxiliary particle sampling technique is 
\begin{itemize}
\item Let $\{(x_{t-1},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t-1}|y_{t-1})$
\end{itemize}
For $k=1,...,N$
\begin{itemize}
\item Draw $I_{k}$ with $P(I_{k}) \propto w_{t-1}^{(i)}f(y_{t}|g(x_{t-1}^{(i)})) $ where $g(x_{t-1}^{(i)})=E(X_{t}|X_{t-1})$
\item Draw $x_{t}^{(k)} \sim N(x_{t-1}^{(I_{k})},\tau^2)$
\item Set  $w_{t}^{(k)} = \frac{f_{N}(y_{t}|x_{t}^{(k)})}{f_{N}(y_{t}|g(x_{t-1}^{(I_{k})}))}$
\end{itemize}
The remaining steps are the same of the particle filter already seen in section XX.

```{r}
APFfun<-function(data,N,m0,C0,tau,sigma,r){
  if(missing(r)){r=2}else{}
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)

  for(t in 1:length(data)){

    weight = w*dnorm(data[t],x,sigma)
    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
    x1   = rnorm(N,x[k],tau)
    lw  = dnorm(data[t],x1,sigma,log=TRUE)-dnorm(data[t],x[k],sigma,log=TRUE)
    w   = exp(lw)
    w   = w/sum(w)
    ESS  = 1/sum(w^2)

    if(ESS<(N/r)){
      index<-sample(N,size=N,replace=T,prob=w)
      x1<-x1[index]
      w<-rep(1/N,N)
    }else{}

    x <- x1
    xs = rbind(xs,x)
    ws = rbind(ws,w)
    ess =rbind(ess,ESS)

  }
  return(list(xs=xs,ws=ws,ess=ess))
}
```

```{r echo=FALSE}
apf<-APFfun(y,N,m0,C0,tau,sigma)
ESSplot(df,apf)
```

```{r echo=FALSE, fig.align='center', fig.cap="Particle Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
Filtplot(df,apf,"Particle Filter")
```
\
\textbf{Liu and West Filter}
\
Consider the toy example of section XX. Let $\psi=(\sigma^{2},\tau^{2})$ be unknown and assign a gamma prior on these parameters
\begin{align*}
\sigma^{2}  & \sim G(\alpha_{v},\beta_{v}) \\
\tau^{2}  & \sim G(\alpha_{w},\beta_{w})
\end{align*}
or let assign them a uniform prior if we have no knowledge on hyperparameters.
After having drown the hyperparameters indipendently form their priors and having set $w_{0}^{(i)}=N^{-1}$, $i=1,...,N$, and $\hat{\pi}_{0}=\sum_{i=1}^{N}w_{0}^{(i)}\delta_{(x_{0}^{(i)},\psi^{(i)})}$, for t=1,..T 
\begin{itemize}
\item Compute $\hat{\psi}=E_{\hat{\pi}_{t-1}}(\psi)$ and $\Sigma=Var_{\hat{\pi}_{t-1}}(\psi)$. For $i=1,...,N$ set
\begin{align*}
m(\psi^{(i)}) & =a\psi^{(i)}+(1-a)\overline{\psi}\\
v(\psi^{(i)}) & =(1-a^2)\Sigma
\end{align*}
and
\begin{align*}
\alpha(\psi^{(i)}) & =\frac{m(\psi^{(i)})^2}{v(\psi^{(i)})} \\
\beta(\psi^{(i)}) & =\frac{m(\psi^{(i)})}{v(\psi^{(i)})}
\end{align*}
For $k=1,...,N$
\begin{itemize}
\item Draw $I_{k}$ with $P(I_{k}=i) \propto w_{t-1^{(i)}}f_{N}(y_{t}|g(x_{t}^{(i)}),m(\psi^{(i)}))$ where for simplicity $g(x_{t}^{(i)})=E(x_{t}|x_{t-1},m(\psi^{(i)}))$
\item Draw $\psi^{(k)} \sim G(\alpha(\psi^{(I_{k})}),\beta(\psi^{(I_{k})}))$
\item Draw $x_{t}^{(k)} \sim N(x_{t-1}^{(I_{k})},\tau^{{2}^{(k)}})$
\item Set $\tilde{w}_{t}^{k}=\frac{f_N(y_{t}|x_{t}^{(k)},\psi=\psi^{(k)})}{f_N(y_{t}|g(x_{t}^{(I_{k})}),\psi=m(\psi)^{(I_k)})}$
\end{itemize}
\item Normalize the weights
\item Compute the effective sample size ($ESS$)
\item If $ESS<N/2$, resample:
\begin{itemize}
\item Draw a sample of size N, $x_{t}^{(1)},...,x_{t}^{(N)}$,from the discrete distribution $P((x_{t},\psi)=(x_{t}^{(i)},\psi^{(i)}))=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{itemize}

\end{itemize}
```{r}
LWfun<-function(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif,r){
  if(missing(r)){r=2}else{}
  xs     = rnorm(N,m0,sqrt(C0))
  if(unif==T){
  pars   = cbind(runif(N,0,10),runif(N,0,10))}else{}
  pars   = cbind(rgamma(N,shape=alphav,scale=betav),rgamma(N,shape=alphaw,scale=betaw))
  a      = (3*delta-1)/(2*delta)
  h2     = 1-a^2
  parss  = array(0,c(N,2,n))
  xss    = NULL
  ws     = NULL
  ess    = NULL
  w      = rep(1/N,N)
  for (t in 1:length(data)){
    meanV = weighted.mean(pars[,1],w)
    varV  = weighted.mean((pars[,1]-meanV)^2,w)
    meanW = weighted.mean(pars[,2],w)
    varW  = weighted.mean((pars[,2]-meanW)^2,w)

    muV = a*pars[,1]+(1-a)*meanV
    sigma2V = (1-a^2)*varV
    alphaV = muV^2/sigma2V
    betaV = muV/sigma2V

    muW = a*pars[,1]+(1-a)*meanW
    sigma2W = (1-a^2)*varW
    alphaW = muW^2/sigma2W
    betaW = muW/sigma2W

    weight      = w*dnorm(data[t],xs,sqrt(muV))
    k           = sample(1:N,size=N,replace=T,prob=weight)

    pars[,1]<-rgamma(N,shape=alphaV[k],rate=betaV[k])
    pars[,2]<-rgamma(N,shape=alphaW[k],rate=betaW[k])

    xsprevious<-xs[k]
    xs = rnorm(N,xs[k],sqrt(pars[,2]))

    w           = exp(dnorm( data[t],xs,sqrt(pars[,1]),log=T)-
                        dnorm( data[t],xsprevious,sqrt(muV[k]),log=T))
    w           = w/sum(w)
    ESS         = 1/sum(w^2)

    if(ESS<(N/r)){
      index<-sample(N,size=N,replace=T,prob=w)
      xs<-xs[index]
      pars<-pars[index,]
      w<-rep(1/N,N)
    }else{
      xs<-xs
      pars<-pars
    }

    xss         = rbind(xss,xs)
    parss[,,t]  = pars 
    ws          = rbind(ws,w)
    ess         = rbind(ess,ESS)
  }
  return(list(xss=xss,parss=parss,ws=ws,ess=ess))
}
```

```{r echo=FALSE}
delta=0.7
lwf<-LWfun(y,N,m0,C0,1,1,1,1,delta,unif=T)
ESSplot(df,lwf)
```

```{r echo=FALSE, fig.align='center', fig.cap="Particle Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
Filtplot(df,lwf,"Liu and West Filter")
```

# COMPARISON (ANOTHER SECTION)
```{r}
library(ggpubr)
library(kableExtra)

Kalmanvarplot<-ggplot(DLM.df,aes(x=timeframe))+
  geom_line(aes(y=C))+
  labs(x="Time",
       y="")+
  ggtitle("Kalman Filter")+
  theme_bw()

Varplot<-function(dataframe,fun,title){

Filt<-Filtervalues(fun)
mean<-Filt$mean
sd<-Filt$sd
dataframe<-data.frame(dataframe,mean,sd)

ggplot(dataframe,aes(x=timeframe))+
geom_line(aes(y=sd))+
  labs(x="Time",
       y="")+
  ggtitle(title)+
  theme_bw()

}

Varsisplot<-Varplot(df,sis,"SIS")
Varpfplot<-Varplot(df,pf,"PF")
Varapfplot<-Varplot(df,pf,"APF")
Varlwplot<-Varplot(df,pf,"LW")

ggarrange(Kalmanvarplot,Varsisplot,Varpfplot,Varapfplot,Varlwplot)

@@

```{r}
truex<-x
estimatedx<-matrix(NA,ncol=6,nrow=6)
colnames(estimatedx)<-c("N","Threshold","KF","PF","APF","LWF")
estimatedx[,1]<-c(20,100,500,1000,1000,1000)
estimatedx[,2]<-c(0.5,0.5,0.5,0.5,0.25,0.1)
RMSE<-function(x,xhat){sqrt(mean((x - xhat)^2))}

i=1
for(N in c(20,100,500)){

DLM1<-DLM(y,sigma2,tau2,m0,C0)
pf1<-PFfun(y,N,m0,C0,tau,sigma)
apf1<-APFfun(y,N,m0,C0,tau,sigma)
lwf1<-LWfun(y,N,m0,C0,1,1,1,1,delta,unif=T)

estimatedx[i,3]<-RMSE(truex,DLM1$m)
estimatedx[i,4]<-RMSE(truex,Filtervalues(pf1)$mean)
estimatedx[i,5]<-RMSE(truex,Filtervalues(apf1)$mean)
estimatedx[i,6]<-RMSE(truex,Filtervalues(lwf1)$mean)
i=i+1

}

i=4
for(k in c(2,5,10)){
  N=1000
DLM2<-DLM(y,sigma2,tau2,m0,C0)
pf2<-PFfun(y,N,m0,C0,tau,sigma,r=k)
apf2<-APFfun(y,N,m0,C0,tau,sigma,r=k)
lwf2<-LWfun(y,N,m0,C0,1,1,1,1,delta,unif=T,r=k)

estimatedx[i,3]<-RMSE(truex,DLM2$m)
estimatedx[i,4]<-RMSE(truex,Filtervalues(pf2)$mean)
estimatedx[i,5]<-RMSE(truex,Filtervalues(apf2)$mean)
estimatedx[i,6]<-RMSE(truex,Filtervalues(lwf2)$mean)
i=i+1

}

```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
kbl(estimatedx[1:3,], align="c", longtable = T, booktabs = T, caption="RMSE",digits=3)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
kbl(estimatedx[4:6,], align="c", longtable = T, booktabs = T, caption="RMSE",digits=3)
```

# STOCHASTIC VOLATILITY

Basic specification of Stochastic Volatilty Model
\begin{align}
y_{t}|x_{t} & \sim N(0,e^{x_{t}}) \\
x_{t}|x_{t-1} & \sim N(\alpha+\beta x_{t-1},\tau^2)
\end{align}
\
```{r include=FALSE}
#############################
#
#   y(t) = exp(h(t)/2)*v(t)
#   x(t) = alpha + beta*x(t-1) + tau*w(t)
#
#   v(t) ~ N(0,1)
#   w(t) ~ N(0,1)
#   x(0) ~ N(0,100)
#
#   PRIOR ON alpha,beta,tau2

# Simulating the data
# -------------------
set.seed(12345)
n     =  50
alpha = -0.0031
beta  =  0.9951
tau2  =  0.0074
tau   = sqrt(tau2)
y     = rep(0,n)
x     = rep(0,n)
x[1]  = alpha/(1-beta)
y[1]  = rnorm(1,0,exp(x[1]/2))
for (t in 2:n){
  x[t] = alpha + beta*x[t-1] + rnorm(1,0,tau)
  y[t] = rnorm(1,0,exp(x[t]/2))
}
alpha.true = alpha
beta.true  = beta
tau2.true  = tau2

timeframe<-c(1:length(y))
dfsv<-data.frame(timeframe,y,x)
```

```{r echo=FALSE}
n=50
tau2=1
sigma2=1
sigma=sqrt(sigma2)
tau=sqrt(tau2)
x0=0

ggplot(dfsv,aes(x=timeframe))+
  geom_line(aes(y=y, linetype="Observations"))+
  geom_line(aes(y=x, linetype="True States"))+
  scale_linetype_manual(name="",
                      values=c("Observations" = 1,
                               "True States" = 3))+
  labs(x="Time",
       y="")+
  ggtitle("States and Observations")+
  theme_bw()+
  theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
        legend.background = element_rect(fill = "white", colour = "gray30")) +
  theme(plot.title = element_text(hjust = 0.5))
```

\textbf{Particle Filter}
\
```{r}
SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
  if(missing(r)){r=2}else{}
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)

  for(t in 1:length(data)){

    x<-rnorm(N,alpha+beta*x,tau)
    w1<-w*dnorm(data[t],0,exp(x/2))

    w = w1/sum(w1)
    ESS  = 1/sum(w^2)

    if(ESS<(N/r)){
      index<-sample(N,size=N,replace=T,prob=w)
      x<-x[index]
      w<-rep(1/N,N)
    }else{}

    xs = rbind(xs,x)
    ws = rbind(ws,w)
    ess =rbind(ess,ESS)
  }
  return(list(xs=xs,ws=ws,ess=ess))
}
```

\textbf{Auxiliary Particle Filter}

```{r}
SVAPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
  if(missing(r)){r=2}else{}
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)

  for(t in 1:length(data)){

    weight = w*dnorm(data[t],0,exp(x/2))
    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
    x1   = rnorm(N,alpha+beta*x[k],tau)
    lw  = dnorm(data[t],0,exp(x/2),log=TRUE)-dnorm(data[t],0,exp(x/2),log=TRUE)
    w   = exp(lw)
    w   = w/sum(w)
    ESS  = 1/sum(w^2)

    if(ESS<(N/r)){
      index<-sample(N,size=N,replace=T,prob=w)
      x1<-x1[index]
      w<-rep(1/N,N)
    }else{}

    x <- x1
    xs = rbind(xs,x)
    ws = rbind(ws,w)
    ess =rbind(ess,ESS)

  }
  return(list(xs=xs,ws=ws,ess=ess))
}
```

\textbf{Liu and West}

```{r include=FALSE}
# Data and prior hyperparameters 
# ------------------------------
m0      = 0.0
C0      = 0.1
sC0     = sqrt(C0)
ealpha  = alpha
valpha  = 0.01
ebeta    = beta
vbeta    = 0.01
nu      = 3
lambda  = tau2

# Liu and West filter
# -------------------
set.seed(8642)
N      = 1000
```

```{r}
SVLWfun<-function(data,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda){

>>>>

---
title: "R Notebook"
output:
  pdf_document: default
  html_notebook: default
---
```{r include=FALSE}
library(ggplot2)
q025=function(x){quantile(x,0.025)}
q975=function(x){quantile(x,0.975)}
```

\section{Implementation}

In this section we present some practical illustrations of the algorithms previously discussed. We start by considering a random walk plus noise model, i.e. the most basic form of a linear Gaussian state-space model. 
\begin{align}
y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \\
x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \\
x_{0} & \sim N(m_{0},C_{0})
\end{align}
As we already know, in this case the filtering distribution can be computed in closed form solutions using the Kalman filter. However, this toy example is useful to understand the logic of the algorithms and compare the performances of different filtering strategies. \
\
\textbf{Kalman Filter}
\
The Kalman Filter for this model can be easily implemented. Starting from the filtering distribution at period $t-1$, $x_{t-1}|y_{1:t-1} \sim N(m_{t-1},C_{t-1})$, we compute:
\begin{itemize}
\item the one-step-ahead predictive distribution at time $t-1$
\begin{align*}
x_{t}|y_{1:t-1} & \sim N(a_{t},R_{t})\\
a_{t} & = m_{t-1}\\
R_{t} & = C_{t-1}+\tau^2
\end{align*}
\item the filtering distribution at time $t$ as $p(x_{t}|y_{1:t}) \propto p(x_{t}|y_{1:t-1})p(y_{t}|x_{t})$
\begin{align*}
x_{t}|y_{1:t} & \sim N(m_{t},C_{t}) \\
m_{t} & = \big(1-\frac{R_{t}}{R_{t}+\sigma^2}\big)a_{t}+\frac{R_{t}}{R_{t}+\sigma^2}y_{t} \\
C_{t} & = \frac{R_{t}}{R_{t}+\sigma^2}\sigma^2
\end{align*}
\end{itemize}
```{r}
DLM<-function(data,sig2,tau2,m0,C0){
  n  = length(data)
  m  = rep(0,n)
  C  = rep(0,n)
  for (t in 1:n){
    if (t==1){
      a = m0
      R = C0 + tau2
    }else{
      a = m[t-1]
      R = C[t-1] + tau2
    }
    A = R/(R+sig2)
    m[t] = (1-A)*a + A*y[t]
    C[t] = A*sig2
  }
  return(list(m=m,C=C))
}
```
\
\textbf{Sequential Importance Sampling}
\
Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \frac{1}{N}\sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. For $t=1,...,n$ where $n$ is the length of the sample, at any iteration
\begin{itemize}
\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
\end{itemize}
```{r}
SISfun<-function(data,N,m0,C0,tau,sigma){
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)
  for(t in 1:length(data)){
    x    = rnorm(N,x,tau)                   #sample from N(x_{t-1},tau)
    w    = w*dnorm(data[t],x,sigma)         #update weight
    xs = rbind(xs,x)
    ws = rbind(ws,w)

@@

```{r}

@@

```{r echo=FALSE, message=FALSE, warning=FALSE}
#Generate a random walk plus noise
#---------------------------------
#n = sample dim
#sigma = obs noise variance
#tau = state noise variance
#x0 = init state

dlm.sim = function(n,sigma,tau,x0){
  x    = rep(0,n)  #state process
  y    = rep(0,n)  #obs process
  x[1] = rnorm(1,x0,tau)
  y[1] = rnorm(1,x[1],sigma)
  for (t in 2:n){
    x[t] = rnorm(1,x[t-1],tau)
    y[t] = rnorm(1,x[t],sigma)
  }
  return(list(x=x,y=y))
}
n=500
tau2=1
sigma2=1
sigma=sqrt(sigma2)
tau=sqrt(tau2)
x0=0
set.seed(1990)
sim<-dlm.sim(n,sigma,tau,x0)
y<-sim$y
x<-sim$x

#Kalman Filter
#-------------
m0=0
C0=100
filtval<-DLM(y,sigma2,tau2,m0,C0)
m<-filtval$m
C<-filtval$C
Low = m + qnorm(0.025)*sqrt(C)
Up = m + qnorm(0.975)*sqrt(C)

#Plots (observetions in black and filtered states in red)
ggplot(DLM.df,aes(x=timeframe))+
  geom_line(aes(y=y))+
  geom_line(aes(y=m),col="red")+
  geom_ribbon(aes(ymin = Low, ymax = Up),
              fill="red",alpha=0.16) +
  labs(x="Periods",
       y="")+
  ggtitle("Kalman Filter")+
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))

@@
